---
title: ""
output:
  html_document:
    highlight: "tango"
#    code_folding: "show"
---

```{r, echo=FALSE, eval=FALSE}
#r chunks are not named
#this is so the Rmd can be called twice in the index.Rmd file
#     once for just the charts
#     once for the charts and the data (don't have to worky about putting into new file )

```

```{r, results="hide", warning=FALSE, message=FALSE}
library(tidyverse)
library(gridExtra)
library(grid)
library(knitr)
library(lubridate)
library(scales)
library(ggrepel)
library(kableExtra)
AccentCol01 <- "firebrick"
knitr::opts_chunk$set(fig.width=10, fig.height=4, dpi=300)
```

The data is taken from [Our World in Data](https://ourworldindata.org/coronavirus) who gets their data from the Eurpoean CDC.  They file can be found [here](https://ourworldindata.org/coronavirus-source-data) and is called `full_data.csv`; also called "All four metrics".   

```{r}
#https://ourworldindata.org/coronavirus-source-data 
covid19 <- read_csv("full_data.csv",
                col_types = cols(
                    date = col_date(format = ""),
                    location = col_character(),
                    new_cases = col_double(),
                    new_deaths = col_double(),
                    total_cases = col_double(),
                    total_deaths = col_double()
                    )
                )
```


```{r, eval=FALSE}
#I'm just look at the data for the United States.   

covid19_colnames <- colnames(covid19)
covid19_US <- covid19 %>%
    mutate(Date = 
           #change column type to date 
           as.Date(
             #parse through current string format mmm d, yyyy
             #requires lubricate package 
             parse_date_time(Date, orders=c("mdy"))
             )
         ) %>% 
  #only look at USA
  filter(Entity == "United States") %>% 
  #only return date and daily count 
  #rename long column names into shorter names 
  select(
      Date
    , "Total Deaths" = covid19_colnames[4]
    , "Total Cases"  = covid19_colnames[5]
    , "Daily Cases"  = covid19_colnames[6]
    )
```

```{r}
covid19_US <- covid19 %>%
  filter(location == "United States")
```


One important calculation is the growth rate.  The growth rate is calculated as follows: 

$$
\text{Growth Rate} = \frac{ \text{New-Cases}_N}{\text{New-Cases}_{N-1}}
$$
where $N$ is a given day.  If the growth rate is below 1 that means the number of new cases is slowing down - this is a positive sign.  

I also wanted to look at the dealth percentage which I calculated as the total number of dealths divided by the total number of cases for a given day.  

```{r}
#number of rows; use for `for loop`  
R <- nrow(covid19_US)

#GROWTH FACTOR  
covid19_US$growth_factor <- NA

#calculate growth rate
#start with 2 because when r=1; 1-1= 0 (don't have a 0th row)
for (r in 2:R){
  covid19_US$`growth_factor`[r] <- 
    #Growth Factor = New Cases Today / New Cases Yesterday 
    covid19_US$new_cases[r] /  covid19_US$new_cases[r-1] 
}


#DEATH PERCENTAGE  
covid19_US$death_percentage <- NA

# calculate dealth percentage  
for (r in 1:R){
  covid19_US$death_percentage[r] <- 
    #dealth percentage = total deaths / total cases 
    covid19_US$total_deaths[r] /  covid19_US$total_cases[r] 
}

```

I am restricted the charts to show values after March 17, 2020 (first date in data set is March 18, 2020).  This is when preventative measures started to increase dramatically.  In addition, given the small number of cases prior to this date caused great variability in the growth rate and has multiple instances where it's  NA or equal to infinity.  

```{r}
startdate <- "2020-03-17"
#3/18 - start 'lockdown' ; after large growth factors↕


covid19_US_adj <- covid19_US  %>%
  #only look at data after a given start date
  filter(date > as.Date(startdate)) %>%
  #remove any NA's that are still present 
  drop_na()
```


```{r}
#find current values to use as labels 

current_date <- max(covid19_US_adj$date)

current_values <- covid19_US_adj %>%
    #return row that has current date
    filter(date == current_date)


current_US_deaths      <- current_values$total_deaths
current_US_totalcases  <- current_values$total_cases
current_US_newcases <- current_values$new_cases
current_US_growthfact  <- current_values$growth_factor
current_US_deathperc   <- current_values$death_percentage
current_US_newdeaths <- current_values$new_deaths
```

```{r}
#set the limits for the x-axis
#want the limits to be the same so the axes for different charts line up
#need to add extra for the bar charts 

x_min <- as.Date(startdate)+.5
x_max <- as.Date(current_date)+1
```

###  Total Cases and Total Deaths  
These show the cumulative total of cases and deaths.  
```{r}
#total cases by day bar chart 

plot_tc <- ggplot(covid19_US_adj, aes(x=date, y=total_cases))+
  #use geom_col because counts per day have alread been counted 
  geom_col(alpha=0.6)+
  ggtitle("Total Cases of COVID-19 in the United States by Day")+
  geom_label(
      data=current_values
    #format label so there is a comma after three places 
    , aes(label = formatC(total_cases, big.mark=",", format="d"))
    )+
  theme_minimal()+
  theme(
     legend.position = "none"
    ,axis.title.x=element_blank()
    ,axis.title.y=element_blank()
  )+
  scale_x_date(breaks=date_breaks("7 days")
              , labels = date_format("%b-%d")
              , minor_breaks=date_breaks("1 day")
              , limits = c(x_min, x_max)  
              )+
  #add commas to numbers in y axis; requires scales package
  scale_y_continuous(labels=comma)
```

```{r}
#total deaths by day bar chart  

plot_td <- ggplot(covid19_US_adj, aes(x=date, y=total_deaths))+
  #use geom_col because counts per day have alread been counted 
  geom_col(alpha=0.6)+
  ggtitle("Total Deaths of COVID-19 in the United States by Day")+
  geom_label(
      data=current_values
    #format label so there is a comma after three places 
    , aes(label = formatC(total_deaths, big.mark=",", format="d"))
    )+
  theme_minimal()+
  theme(
     legend.position = "none"
    ,axis.title.x=element_blank()
    ,axis.title.y=element_blank()
  )+
  scale_x_date(breaks=date_breaks("7 days")
              , labels = date_format("%b-%d")
              , minor_breaks=date_breaks("1 day")
              , limits = c(x_min, x_max)    
              )+
  #add commas to numbers in y axis; requires scales package
  scale_y_continuous(labels=comma)
```

```{r, fig.height=8}  
#Put the plots together and have the plot axes line up
gTC <- ggplotGrob(plot_tc)
gTD <- ggplotGrob(plot_td)
grid.newpage()
grid.draw(rbind(gTC, gTD))
```

### Growth Factor and New Cases 
In order to see if the pandemic is slowing down, it is helpful to look at the growth factor and the daily new cases.  For both these values; we want to see a decrease in values from day to day - which indicates that less people are getting sick.  Recall that a growth factor less than 1 is a positive sign that the growth of cases is slowing down.  Growth factor is directly related to new cases - if daily new cases goes now the growth factor is less than 1; if daily cases goes up the growth factor is greater than one.  These two charts represent the same information in two differnt ways.  



```{r}
#growth factor by date line chart 

caption <- "A growth factor above 1 means new cases are increasing"

plot_gf  <- ggplot(covid19_US_adj, aes(x=date, y=growth_factor))+
  #add red box to show growth rates that are above 1
  annotate(
      "rect"
    , xmin = x_min
    , xmax = x_max
    , ymin =  1
    , ymax =  Inf
    , fill = AccentCol01
    , alpha = 0.2
    ) +
  #add line at 1 
  annotate(
      "segment"
    , x     = x_min
    , xend  = x_max
    , y     =  1
    , yend  =  1
    , color = AccentCol01
    , linetype = "dotted"
    , size = 1
    ) +
  geom_point(size=2, alpha=0.6)+
  geom_line(size=1, alpha=0.6)+
 # geom_hline(yintercept = 1, linetype="dotted", color=AccentCol01, size=1)+
  ggtitle("Growth Factor of COVID-19 in the United States by Day")+
  # ADD LABEL 
  geom_label(
      data = current_values
    , aes(label = round(growth_factor, 4))
    , vjust=0
    , nudge_y = 0.1
    )+
  theme_minimal()+
  theme(
     legend.position = "none"
    ,axis.title.x=element_blank()
    ,axis.title.y=element_blank()
  )+
  scale_x_date(breaks=date_breaks("7 days")
              , labels = date_format("%b-%d")
              , minor_breaks=date_breaks("1 day")
              , limits = c(x_min, x_max) 
              )+
  #want growth rate to go to 0
  ylim(0, NA)+
  #add text 
  annotate("text"
           , x = x_max - 1
           , y = 2
           , label = caption
           , vjust="inward"
           , hjust= 1
           )
```

```{r}  
#new cases by date bar chart    

plot_dc <- ggplot(covid19_US_adj, aes(x=date, y=new_cases))+
  #use geom_col because counts per day have alread been counted 
  geom_col(alpha=0.6)+
  ggtitle("New Cases of COVID-19 in the United States by Day")+
  # ADD LABEL 
  geom_label(
      data = current_values
    #format label so there is a comma after three places 
    , aes(label = formatC(new_cases, big.mark=",", format="d"))
    , vjust = 0
    )+
  theme_minimal()+
  theme(
    #remove all legents 
     legend.position = "none"
     # remove x and y axes 
    , axis.title.x = element_blank()
    , axis.title.y = element_blank()
  )+
  scale_x_date(breaks=date_breaks("7 days")
              , labels = date_format("%b-%d")
              , minor_breaks=date_breaks("1 day")
              , limits = c(x_min, x_max)   
              )+
  #add commas to numbers in y axis; requires scales package
  scale_y_continuous(labels=comma)
```

```{r, fig.height=8}  
#Put the plots together and have the plot axes line up

gGF <- ggplotGrob(plot_gf)
gDC <- ggplotGrob(plot_dc)
grid.newpage()
grid.draw(rbind(gGF, gDC))
```

### New Deaths and Death Percentage   

```{r}
#start by defining two fucntion

#round number to specific base  
func_mround <- function(x, base){
  base*round(x/base)
}

#round up to a certain base; this will be used for upper bound for primary axis 
Ylim <- function(curr_val, base){
round_curr_val <- func_mround(curr_val, base)
if( round_curr_val - curr_val < 0 ){
  lim <- round_curr_val + base
} else {
  lim <- round_curr_val
}
return(lim)
}
```




```{r}
#td = total deaths 
#set base for primary axis 
dd_base <- 1000

#primary/main y-axis limit 
dd_y1lim  <- Ylim(max(covid19_US_adj$new_deaths), dd_base) + dd_base
  #add an extra base to add more room 

#max value for the secondary axis 
maxdeathperc <- max(covid19_US_adj$death_percentage, na.rm=TRUE)

#TRANSFORM PRIMARY AXIS TO SECONDARY AXIS 
#in order to have a secondary axis have to have a multiplier to transform primary axis to sec axis
#find the Ylim of primary axis
# next want to find an upper bound for seconday axis
#     want both upper bounds to be divisible by the same number 
#     i.e. if the primary axis is from 0 to 20 and it has 4 breaks (5, 10, 15, 20)
#     I want to set the upper bound of the secondary axis to a value divisible by 4
#     want teh break lines of the primary axis to match up with the breaks of the secondary axis 
#what if I don't do this?
#     the text breaks of the seconday axis will not line up with the line breaks of the primary axis 

#find what the divisor is 
divisor <- dd_y1lim/dd_base

# use divisor to find acceptable base for sec axis 
dp_base <- divisor/100

#find rounded value of sec axis with sec base 
Rmaxdeathperc <- func_mround(maxdeathperc, dp_base)

#round sec-axis value up; for the sec axis Ylimit
if( Rmaxdeathperc - maxdeathperc < 0 ){
  dp_y2lim <- Rmaxdeathperc + dp_base
} else {
  dp_y2lim <- Rmaxdeathperc
}

#calculate the transform variable  
dd_transform <- dd_y1lim*(1/dp_y2lim)



```

```{r}
#plot the primary data: new deaths by day bar chart    

plot_dd <- ggplot(covid19_US_adj, aes(x=date))+
  #use geom_col because counts per day have alread been counted 
  geom_col(aes(y=new_deaths), alpha=0.6)+
  #add total dealths label 
  geom_label(
      data = current_values
    #format label so there is a comma after three places 
    , aes(
        label = formatC(new_deaths, big.mark=",", format="d")
      , y     = new_deaths
      )
    )
```

```{r}
#Then, plot the secondary data: death percentage.  

plot_dd <- plot_dd + 
  geom_line(
    data = covid19_US_adj, 
    #transformatio to match secondary axis below
      aes(y=death_percentage*dd_transform)
    , color=AccentCol01
    , size= 1
    )+
  geom_point(
    data = covid19_US_adj, 
    #transformatio to match secondary axis below
      aes(y=death_percentage*dd_transform)
    #color 
    , color=AccentCol01
    , size= 2
    )+
  #add label to most recent value 
  geom_label(
      data = current_values
    , aes(
        label = percent(death_percentage, accuracy=0.001)
      # remember to add the transform! 
      , y     = death_percentage*dd_transform 
      )
    , hjust = 0.25
    , vjust = 0
    , color = AccentCol01
    )

```

```{r }
#It is imparative to add the scales.  

plot_dd <- plot_dd + 
  #scale the date to be onece a week 
  #add specific limits so it will line up with geom_line 
  scale_x_date(breaks=date_breaks("7 days")
              , labels = date_format("%b-%d")
              , minor_breaks=date_breaks("1 day")
              , limits = c(x_min, x_max)    
              )+
  #add commas to numbers in y axis; requires scales package
  scale_y_continuous(
      limits = c(0, dd_y1lim)
    #add commas to numbers in y axis; requires scales package
    , labels = comma
    #this transformation MUST match the transformation above
    #divide by current_US_deaths leads to an axis of 0 to 1
    #divide by current_US_deaths*10 leads to an axis of 0 to 0.1
    , sec.axis = sec_axis(~./dd_transform)
    )
```

```{r }
#Finally, a title and other themes are specified.    

plot_dd <- plot_dd + 
  #add title 
  ggtitle("New Deaths and Dealth Percentage of COVID-19 in the United States by Day")+
  #minimal theme
  theme_minimal()+
  theme(
    #remove all legents 
     legend.position = "none"
     # remove x and y axes 
    , axis.title.x = element_blank()
    , axis.title.y = element_blank()
  )

plot_dd
```

#### Summary  

Although the growth rate appears to be starting to slow down; the death percentage has been consistently going up over the past 4 weeks and is almost at 5%.  

### Values for Past Week 

```{r}
#transpose table; easier to read
covid19_US_adj %>%
  arrange(desc(date)) %>%
  top_n(7, wt=date) %>%
  #format data for output tabe
  mutate(
      Date = format(date, '%a, %b %d, %Y')
    , `Total Cases` = formatC(total_cases, big.mark=",", format="d")
    , `New Cases`  = formatC(new_cases, big.mark=",", format="d")
    , `Growth Factor` = round(growth_factor, 4)
    , `Total Deaths` = formatC(total_deaths, big.mark=",", format="d")
    , `New Deaths` = formatC(new_deaths, big.mark=",", format="d")
    , `Death Percentage` = percent(death_percentage, accuracy=0.001)
  ) %>%
  #re-order to group cases and death information together 
  select(
      Date
    , `Total Cases`
    , `New Cases`
    , `Growth Factor`
    , `Total Deaths`
    , `New Deaths`
    , `Death Percentage`
    ) %>%
  kable(align=c('l', rep('r', 5))) %>%
  kable_styling(
    bootstrap_options = c(
      #adds stiped color to rows
        "striped"
      #highlight row when hover over it 
      , "hover"
      #table doesn't have to be full width 
      , full_width = FALSE
      #make it horizontally scrollable
      , "responsive"
      )
    ) %>%
  row_spec(1, bold=T, color="white", background = "rgba(0, 0, 0, .5)")

```
