---
title: "Tracking COVID-19 in the United States"
output:
  html_document:
    highlight: "tango"
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo=FALSE) 
```


```{r index-outside-scripts, results="hide", warning=FALSE, message=FALSE, include = FALSE}
source("script/variable/setup.R")
source("script/dataframe/covid19_state.R") #run script for covid19 df
source("script/dataframe/covid19_county.R") #run script for covid19 df
source("script/variable/parameters.R") #global parameters
```




```{r index-mod-dates}
mod_dates <- tribble(
       ~type,           ~name, ~mod_date
  , "script", "setup"        , file.info("script/variable/setup.R")$mtime
  , "script", "colors"       , file.info("script/variable/colors.R")$mtime
  , "script", "covid df"     , file.info("script/dataframe/covid19_state.R")$mtime
  , "script", "covid cnty df", file.info("script/dataframe/covid19_county.R")$mtime
  , "script", "param"        , file.info("script/variable/parameters.R")$mtime
  , "rmd"   , "overview"     , file.info("overview.Rmd")$mtime
  , "rmd"   , "growth_factor", file.info("growth_factor.Rmd")$mtime
  , "rmd"   , "state"        , file.info("state.Rmd")$mtime
  , "rmd"   , "data_source"  , file.info("data_source.Rmd")$mtime
  , "rmd"   , "nc"      , file.info("nc.Rmd")$mtime
  , "rmd"   , "total"        , file.info("total_cases.Rmd")$mtime
)


lastmod_date <- mod_dates %>% pull(mod_date) %>% max() %>% with_tz(tzone = "America/Los_Angeles") %>% format("%B, %e %Y %H:%M %Z")
data_date <- covid19 %>% pull(date) %>% max() %>% format("%B, %e %Y")
```


# {.tabset}


## Overview  

```{r child="overview.Rmd"}
```
<br><br>
<center>
JH data as of `r data_date `   <br>
<a href="https://github.com/MarEichler/covid19" target = "_blank">
Code </a>
last updated `r lastmod_date`
</center>


## Growth Factor  

```{r child="growth_factor.Rmd"}
```
<br><br>
<center>
JH data as of `r data_date `   <br>
<a href="https://github.com/MarEichler/covid19" target = "_blank">
Code </a>
last updated `r lastmod_date`
</center>

## State Trends

```{r child="state.Rmd"}
```
<br><br>
<center>
JH data as of `r data_date `   <br>
<a href="https://github.com/MarEichler/covid19" target = "_blank">
Code </a>
last updated `r lastmod_date`
</center>


## Current New Cases

```{r child="nc.Rmd"}
```
<br><br>
<center>
JH data as of `r data_date `   <br>
<a href="https://github.com/MarEichler/covid19" target = "_blank">
Code </a>
last updated `r lastmod_date`
</center>


## Total Cases

```{r child="total_cases.Rmd"}
```

<br><br>
<center>
JH data as of `r data_date `   <br>
<a href="https://github.com/MarEichler/covid19" target = "_blank">
Code </a>
last updated `r lastmod_date`
</center>

## Data  
```{r child="data_source.Rmd"}
```


<br><br>
<center>
JH data as of `r data_date `   <br>
<a href="https://github.com/MarEichler/covid19" target = "_blank">
Code </a>
last updated `r lastmod_date`
</center>