---
title: ""
output:
  html_document:
    highlight: "tango"
editor_options: 
  chunk_output_type: inline
---


```{r packages-global-options, results="hide", warning=FALSE, message=FALSE}
library(tidyverse)
library(gridExtra)
library(grid)
library(knitr)
library(lubridate)
library(scales)
library(ggrepel)
library(kableExtra)
library(zoo)
knitr::opts_chunk$set(fig.width=8, fig.height = 4, dpi=300)
```

```{r import-colors}
#read in colors 
source("script/colors.R")
```


This data is downloaded from 
<a href="https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/" target="_blank" >
USA Facts CDC</a>. This data requires formating the dates and calculating the new cases and deaths by day.  USA Facts gets data by county on a daily basis, this is totaled to get values for each day for the entire US. 

The
<a href="https://www.cdc.gov/coronavirus/2019-ncov/cases-updates/cases-in-us.html" target="_blank">
American CDC </a> 
links to USA Facts under Cases & Death by County.  Although for the below maps data per state/county is not used, information on that level is used on the next tab when ploting maps.    


```{r download-process-data}
source("script/usafacts_data.R")
```


```{r load-wrangle-data}
load("data/covid19_US.rda")
#number of rows; use in `for loop`  
R <- nrow(covid19_US)

#GROWTH FACTOR  
covid19_US$growth_factor <- NA

#calculate growth rate
#start with 2 because when r=1; 1-1= 0 (don't have a 0th row)
for (r in 2:R){
  covid19_US$growth_factor[r] <- 
    #Growth Factor = New Cases Today / New Cases Yesterday 
    covid19_US$new_cases[r] /  covid19_US$new_cases[r-1] 
}

covid19_US <- covid19_US %>%
  mutate(
    growth_factor = ifelse(growth_factor == Inf, new_cases, growth_factor)
         )

#DEATH PERCENTAGE  
covid19_US <- mutate(covid19_US, death_percentage = total_deaths / total_cases)

#calculate moving average 
ma_k <- 14
covid19_US$MA_growth_factor <- c(rep(NA, ma_k - 1), rollmean(covid19_US$growth_factor, k=ma_k))
covid19_US$MA_new_cases <- c(rep(NA, ma_k - 1), rollmean(covid19_US$new_cases, k=ma_k))


#save copy of data 
write.csv(covid19_US, "data/data_usafacts.csv")
```


I am restricted the charts to show values on March 15, 2020 and after.  This is when preventative measures started to increase dramatically.  In addition, given the small number of cases prior to this date caused great variability in the growth rate and has multiple instances where it's  NA or equal to infinity.  

```{r start-date}
startdate <- "2020-03-15"
#3/18 - start 'lockdown' ; after large growth factors↕

covid19_US_adj <- covid19_US  %>%
  #only look at data after a given start date
  filter(date >= as.Date(startdate))
```


```{r current-values}
#find current values to use as labels 

current_date <- max(covid19_US_adj$date)

current_values <- covid19_US_adj %>%
    #return row that has current date
    filter(date == current_date)


current_US_deaths      <- current_values$total_deaths
current_US_totalcases  <- current_values$total_cases
current_US_newcases <- current_values$new_cases
current_US_growthfact  <- current_values$growth_factor
current_US_deathperc   <- current_values$death_percentage
current_US_newdeaths <- current_values$new_deaths
```

 

```{r gen-scaling}
#set the limits for the x-axis
#want the limits to be the same so the axes for different charts line up
#need to add extra for the bar charts 

x_min <- as.Date(startdate)-.5
x_max <- as.Date(current_date)+1

#ggplot set ups 
date_scaling <- scale_x_date(
      name = NULL
    , labels = date_format("%b-%d")
    , limits = c(x_min, x_max)
    , expand = c(0, 0)
  )

thousands_scaling <- scale_y_continuous(
      name = NULL
    , limits = c(0, NA)
    , labels = scales::unit_format(unit = "k", scale = 1e-3, accuracy = 1, big.mark = ",")
    , expand = c(0, 0)
  )

title_size <- 14
subtitle_size <- 12
```

###  Total Cases and Total Deaths  
These show the cumulative total of cases and deaths.  These total values are important; however they are not helpful for figuring out whether the pandemic is slowing down or growing.  

```{r tc-td-plot, fig.height=3}
covid_tidy <- covid19_US_adj %>%
  select(date, total_cases, total_deaths) %>%
  pivot_longer(
      cols = c(total_cases, total_deaths)
    , names_to = "type"
    , values_to = "count"
  ) 

total_labels <- covid_tidy %>%
  filter(date == current_date) %>%
  mutate(
    label = paste("Current Total", formatC(count, big.mark=",", format="d"), sep = "\n")
  )

ggplot(covid_tidy, aes(date, count)) + 
  geom_area(alpha = 0.4)+
  geom_line(
      size = 1
    , alpha = 0.6
  )+
  facet_wrap(
    .~factor(type, labels = c("Total Cases by Day", "Total Dealths by Day"))
    , ncol=2
    , scales = "free_y"
  ) + 
  date_scaling +
  thousands_scaling +
  geom_text(
      data = total_labels 
    #format label so there is a comma after three places 
    , aes(
          label = label
        , y = -Inf
        , x = current_date
      )
    , vjust = -0.15
    , hjust=1.05 
    , size = 6
    , fontface = "bold"
    , color = "white"
    ) +
  theme_minimal() +
#  ggtitle("COVID-19 in the United States by Day") +
  theme(
      plot.title = element_text(size = title_size, hjust = 0.5)
  #  , axis.text.x = element_text(hjust = 0.2)
    , strip.text.x = element_text(size = subtitle_size)
    , panel.spacing = unit(.5, "in")
    , plot.margin  = margin(t=0, r=15, b=0, l=0)
    , axis.ticks.length = unit(5, "pt")
  )

```


### New Cases 

Looking at new cases each day can help us see if the pandemic is slowing (new cases per day is decreasing). 


```{r dc-plot-setup}  
#daily cases set up 

plot_dc <- ggplot(covid19_US_adj, aes(x=date, y=new_cases))+
  geom_col(alpha=0.3) +
  date_scaling + 
  thousands_scaling +
  theme_minimal()+
  theme(
      plot.title = element_text(hjust = 0.5, size = subtitle_size)
    , axis.title.x=element_blank()
    , axis.title.y=element_blank()
    , plot.margin  = margin(t=0, r=15, b=0, l=0)
    , axis.ticks.length = unit(5, "pt")
  )
```


There can be a lot of variability in the daily points - this is due to many different variables.  On variable is the availability of tests, cases will go down if there is scarcity of tests and rise dramatically when more tests become available.  One way to help get a better sense over the overall trend is by smoothing the data.  


```{r loess-plots, warning = FALSE}  
#LOESS CURVE

title_l <- "LOESS Smoothing "

loess_smooth <- stat_smooth(
      geom="line"
    , size = 1.25
    , color = blue_comp
    , se = FALSE 
    , alpha = 0.6
    , method = "loess"
    , formula = y~x
    , span = 0.9
  )



plot_dc_l <- plot_dc +
  loess_smooth +
  ggtitle(title_l) 


title_ma <- paste("Moving Average, k =", ma_k)


plot_dc_ma <- plot_dc +
  geom_line(
    aes(y = MA_new_cases)
    , size = 1.25
    , color = blue_comp
    , alpha = 0.6
  ) +
  ggtitle(title_ma)

grid.arrange(
    plot_dc_l
  , plot_dc_ma
  , nrow = 1
  ,top=textGrob(
      paste("New Cases per Day", sep = "\n")
    , gp=gpar(fontsize=title_size, fontface = "bold")
  )
)
```


### New Deaths and Death Percentage   

I also wanted to look at the dealth percentage which I calculated as the total number of deaths divided by the total number of cases for a given day.  


```{r dp-plot}
plot_dp <- ggplot(covid19_US_adj, aes(date, death_percentage)) + 
  geom_line(color = "grey40", size = 1) + 
  date_scaling + 
  scale_y_continuous(
      name = "Total Dealths / Total Cases"
    , limits = c(0, NA)
    , labels = scales::percent
  ) +
  ggtitle("Dealth Percentage per Day") + 
  theme_minimal()+
  theme(
      plot.title = element_text(hjust = 0.5, size = subtitle_size)
    , axis.title.x=element_blank()
    , plot.margin  = margin(t=0, r=15, b=15, l=0)
    , axis.ticks.length = unit(5, "pt")
  )
```

```{r nd-plot}
plot_nd <- ggplot(covid19_US_adj, aes(date, new_deaths)) + 
  geom_col(alpha=0.3) +
  date_scaling + 
  thousands_scaling +
  ggtitle("New Deaths per Day") + 
  theme_minimal()+
  theme(
      plot.title = element_text(hjust = 0.5, size = subtitle_size)
    , axis.title.x=element_blank()
    , axis.title.y=element_blank()
    , plot.margin  = margin(t=15, r=15, b=0, l=0)
    , axis.ticks.length = unit(5, "pt")
  )

```


```{r dp-nd-plot}
cowplot::plot_grid(
    plot_dp
  , plot_nd
  , nrow = 1
  , align = "h"
)
```





### Values for Past 14 Days 

```{r values-table}
#transpose table; easier to read
covid19_US_adj %>%
  arrange(desc(date)) %>%
  top_n(14, wt=date) %>%
  #format data for output tabe
  mutate(
      Date = format(date, '%a, %b %d, %Y')
    , `Total Cases` = formatC(total_cases, big.mark=",", format="d")
    , `New Cases`  = formatC(new_cases, big.mark=",", format="d")
    , `Total Deaths` = formatC(total_deaths, big.mark=",", format="d")
    , `New Deaths` = formatC(new_deaths, big.mark=",", format="d")
    , `Death Percentage` = percent(death_percentage, accuracy=0.001)
  ) %>%
  #re-order to group cases and death information together 
  select(
      Date
    , `Total Cases`
    , `New Cases`
    , `Total Deaths`
    , `New Deaths`
    , `Death Percentage`
    ) %>%
  kable(
    align=c('l', rep('r', 5))
    , escape = F
  ) %>%
  kable_styling(
    bootstrap_options = c(
      #adds stiped color to rows
        "striped"
      #highlight row when hover over it 
      , "hover"
      #table doesn't have to be full width 
      , full_width = FALSE
      #make it horizontally scrollable
      , "responsive"
      )
    ) %>%
  row_spec(1, bold=T, color="white", background = "rgba(0, 0, 0, .5)")
```
